Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses

### 2014-10-17 paper_v1.Rnw
integrating RQs and Hs into lit. review

Research and theorizing on attrition in education has a rich history. This review is intended to serve as a foundation to build on with the current research. The focus of this breif review is on how ways of thinking about attrition have developed over the last decades.

@@

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college. While piror experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration. Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively. This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2011brief}.

Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration. Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.

@@

A survey of over one hundred learners who dropped out of MOOC showed that most learners indicated too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}. A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}. Based on just over one thousand learners who participated in the discussion forum of a particular MOOC, the likelihood of dropout was lower for those who actively participated in the first week of the course, those who served as an authority figure in the community on the forum, and those who did not engage in a particular subcommunity on the forum \cite{rose2014social}.

A number of machine learning approaches to predicting attrition have been tested, incluging hidden Markov models \cite{balakrishnan2013predicting},

@@

\begin{description}
  \item[RQ1] Why do people who drop out report they drop out?
  \item[RQ2] Who is most likely to drop out in terms of demographics, intentions/motivations, geo location, prior experience with online courses?
  \item[RQ3] What reasons do people from different subgroups (demographic, intention, etc.) report for dropping out?
  \item[RQ4] What proportion of people who dropped out were satisfied with what they got out of the course?
  \item[RQ5] What behaviors do people with different dropout reasons exhibit?
  \item[H1] prior experience -- technical difficulties
  \item[H2] demographics (education) -- course difficulty
\end{description}

@@

H2: people who report external committments (family, school), are more likely to specify the reason than people who have `no time' because they have low volition

>>>>

Research and theorizing on attrition in education has a rich history. This review is intended to serve as a foundation to build on with the current research. We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work. The focus of this breif review is on how ways of thinking about attrition have developed over the last decades.

@@

\begin{description}
  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC?
  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
\end{description}

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college. While piror experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration. Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively. This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2007question}.

\begin{description}
  \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
\end{description}

Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration. Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}. Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (i.e., instilling a growth mindset) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}. The results highlight the powerful influence of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.

\begin{description}
  \item[RQ2] What proportion of learners who dropped out were satisfied with their achievements in the course?
  \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
\end{description}

@@

A survey of over one hundred learners who dropped out of a MOOC showed that the majority indicated having too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}. A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}. A survival analysis of close to 800 learners who had posted on a MOOC discussion forum suggested that the likelihood of dropout was lower for those who i. actively participated in the first week of the course; ii. served as an authority figure in the community on the forum, and iii. did not engage in a particular subcommunity on the forum \cite{yang2013turn}. In addition, a number of machine learning approaches yielded promising results for early dropout prediction (e.g. \cite{taylor2014likely,halawa2014dropout}), but fell short of offering insights into why a given learner would leave the course.

\begin{description}
\item[RQ3a] What reasons do learners report for dropping out of a MOOC?
\item[RQ3b] How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)?
\item[RQ4] How does behavior in MOOCs vary by reasons for dropping out?
\end{description}

@@

% \begin{description}
%   \item[H1] prior experience -- technical difficulties
%   \item[H2] demographics (education) -- course difficulty
% \end{description}

@@

H2: people who report external committments (family, school), are more likely to specify the reason than people who have `no time' because they have low volition.
